---
title: "Demonstrating `clusterProfiler` for term enrichment"
author: "Darach"
date: "`r Sys.Date()`"
---

To use this, you're going to need to install the libraries
`clusterProfiler` and `org.Sc.sgd.db` from Bioconductor, using
their own package management functions. I also make use of `tidyverse`
because it's handy, and I'm assuming you're familiar with it.

```{r,libs,cache=F,warning=F,message=F}
library(tidyverse)
```

# Preliminaries

Here, we read in a few tables. These were downloaded from SGD,
so the GO slim terms are from 
[https://downloads.yeastgenome.org/curation/literature/go_slim_mapping.tab](
https://downloads.yeastgenome.org/curation/literature/go_slim_mapping.tab)
and the rest are likely from 
[http://www.geneontology.org/page/download-go-annotations](
http://www.geneontology.org/page/download-go-annotations).

Then, I make a tibble that is handy for later mappings.

```{r,getGOtermz,cache=T,message=F}
# GOSlim annotation table
read_tsv("data/sgd_go_slim_171013.txt",col_names=F,
  comment="!") -> SGDGOSlim

# Full GO annotation table
read_tsv("data/sgd_go_full_171013.txt",col_names=F,
  comment="!") -> SGDGOFull

# Go terms look up table
read_tsv("data/sgd_go_terms_171013.txt",col_names=F,
  comment="!") %>%
  mutate(GOID=str_c("GO:"
    ,str_pad(string=X1,width=7,side="left",pad="0"))
    ) -> SGDGOTerms

# This is a handy table for mapping systematic names to SGD ID
SystematicToSGDIDmapper <- SGDGOSlim%>%
  select(X1,X2,X3)%>%
  rename(Systematic=X1,SGDID=X3,Common=X2)%>%
  unique()
```


Here's the data. It's from the Barseq after FACS after FISH, aka
the Great Red Herring Fishing Expedition of 2015-2017. 

```{r,dataToAnalyze,cache=T,message=F,warning=F}
datar <- read_csv("data/Figure4_Table_BFFfilteredPooledModels.csv",
  comment="#",
  col_names=c("Systematic","Common","PreShift_Mean",
    "PostShift_Mean","Change_Mean","ResidualFromExpected_Mean"))
```

The data has a few columns. Namely, there's the mean signal before
and after a shift. What does this sorta look like? Well, we could
look at the distribution of this mean after the shift for all
strains, and draw a line at 7.5 for reasons that will become clear.

```{r,histogram,cache=T,warning=F}
ggplot(datar)+aes(x=PostShift_Mean)+geom_histogram(bins=50)+
  geom_vline(xintercept=7.5)
```

# Categorical GO, using the Bioconductor database package

We could do a categorical test of what's above that 7.5 line.
That's fast to compute. For this, I'm going
to use the `bitr` function from `clusterProfiler`.
Note that I'm using the double colon notation to avoid loading
the package, because it carries with it a bunch of `bioconductor`
cruft.

First, we make two lists of IDs, just separating
on the mean being more than 7.5, 
and making a query and background set.

```{r,clusterProfilerSetupCateogries,cache=T,message=F}
TopPostShift_Entrez <- datar%>% filter(PostShift_Mean>7.5)%>%
  left_join(SystematicToSGDIDmapper,by="Systematic")%>%
  {clusterProfiler::bitr(.$SGDID,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,drop=T)} %>%
  pull(ENTREZID)

Background_Entrez <- datar%>% 
  left_join(SystematicToSGDIDmapper,by="Systematic")%>%
  {clusterProfiler::bitr(.$SGDID,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,drop=T)} %>%
  pull(ENTREZID)

head(TopPostShift_Entrez)
head(Background_Entrez)
```

Note that these are now lists of ENTREZ IDs. Note that some of the
terms didn't translate over, so we're missing those. What are those?

```{r,clusterProfilerLookAtFailures,cache=T,message=F}
Background_SGD <- datar%>% 
  left_join(SystematicToSGDIDmapper,by="Systematic")%>%
  {clusterProfiler::bitr(.$SGDID,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,drop=T)} %>%
  pull(SGD)

fail_to_map <- datar%>% 
  left_join(SystematicToSGDIDmapper,by="Systematic")%>%
  {clusterProfiler::bitr(.$SGDID,fromType="SGD",toType="ENTREZID"
    ,OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,drop=T)} %>% as_tibble() %>%
  {anti_join(x=SystematicToSGDIDmapper%>%rename(SGD=SGDID),.,by="SGD")} 

Background_SGD %in% fail_to_map$SGD %>% any()

head(data.frame(fail_to_map),20)
tail(data.frame(fail_to_map),20)
```

The would appear to be snRNA, tRNA, etc. Odd that they're not 
annotated. If this is important for you, you can also introduce
custom annotation lists in `clusterProfiler`. Note that this is
based on the Bioconductor package for annotation, so unless you're
sure it's current for your research, you might want to double check
it's current.

Let's test these for each category of GO terms, look for enrichment.

```{r,categoricalGOenrichment,cache=T}
CategoricalGO <- list()

for (i in c("MF","BP","CC")) {
  CategoricalGO[[i]] <- clusterProfiler::enrichGO(
      gene=TopPostShift_Entrez,
      universe=Background_Entrez,
      OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,
      ont=i,
      pAdjustMethod="fdr",
      qvalueCutoff=0.20,
      minGSSize=2,maxGSSize=200)@result%>%
    as_tibble()%>%mutate(Type=i)
}
```

At each step, this looks for the enrichment of each type of ontology
in that set. Then, we'll push that around into one big table,
and then we're going to do some `tidy`/`purrr` stuff with the `map`
calls to get the common names back again.

```{r,categoricalGOtable,cache=T}
categoricalGOtable <- bind_rows(CategoricalGO) %>% as_tibble %>% 
    # First we bind rows and make it a tibble
  select(Type,ID,Description,qvalue,
    Count,geneID) %>%
    # Then we pick the variables we want
  arrange(Type,-qvalue) %>%
  mutate(geneIDz=map(geneID,str_split,"/",simplify=T))%>%
  mutate(commonz=map(geneIDz,function(x){
    clusterProfiler::bitr(x,toType="SGD",fromType="ENTREZID"
      ,OrgDb=org.Sc.sgd.db::org.Sc.sgd.db,drop=T) %>% as_tibble() %>%
      left_join(SystematicToSGDIDmapper%>%rename(SGD=SGDID),by="SGD")%>%
      pull(Common)
    }))%>%
  select(-geneID,-geneIDz)%>%
  mutate(commonz=unlist(map(commonz,str_c,collapse=",")))
```

```{r,categoricalGOprint,cache=F}
write_lines("#
# Supplementary Table
#
# Results of that categorical GO analysis. Put a description
# here please.
#
# Key:
# Type of enrichment , GO term ID , GO term , q-value , 
#   size of GO term count , common names driving this",
  path="output_table_categorical.csv")

categoricalGOtable %>% 
  write_csv(path="output_table_categorical.csv",append=T)
```

# GSEA using custom terms

Okay, but what about GSEA?

So, let's sort the names by that variables, and then see who's 
enriched. 


First we do one, as a stripped down example. Then we move to doing
lots in a loop. 
For both, we're going to make our sorted gene list:

```{r,sortedGeneList,cache=T}
# First, we make a named list of values, and sort by it
this_gene_list <- 
  datar[order(datar[["PostShift_Mean"]],decreasing=T),] %>%
    # First, it is sorted by a variable. 
    # This non-tidy evaluation is used 
    # because I could not find an easy way to evaluate 
    # it programmatically within tidverse's NSE choice.
  left_join(SystematicToSGDIDmapper,by="Systematic")%>%
    # Then it is joined to a mapper table defined above, to map
    # to the SGD ID
  {setNames(.[["PostShift_Mean"]],nm=.[["SGDID"]])}%>%na.omit()%>%c()
    # Then this is used to name the vector of values of 
    # PostShift_Mean as the SGD ID's. This is na.omited, then c'd
    # into a plain named vector.

head(this_gene_list)
```

```{r,single,dependson="sortedGeneList",cache=T}
singleRun <- clusterProfiler::GSEA(
  geneList=this_gene_list,
    # This is then piped into GSEA function as the gene list
  TERM2GENE=SGDGOSlim[,c("X6","X3")],
    # This provides a mapping of term and gene ID (SGDID)
  TERM2NAME=SGDGOTerms[,c("GOID","X2")],
    # This provides a mapping of term ID to term description
  minGSSize=3,maxGSSize=500,
    # This controls the tested gene set sizes
  pAdjustMethod="fdr",
    # This sets the method of p-value adjusting
  pvalueCutoff=0.5,
    # This sets the p-value cut off
  nPerm=1e4,seed="seedy"
    # This sets the permutations for the boostrapped p-value, and
    # the random seed
  )@result %>% as_tibble() 

head(singleRun)
```

At the end there, it's pulled out using the `@result` accessor from
the return object, then made into a tibble. I forgot what the `@` 
thing is called, but it's like a `$`, but fundamentally different
in a way that doesn't impact us lay user.

I set up a parameter of number of permutations. 
`1e6` takes *forever* to run. So test it out with high p.adjust and
low permutes, then move it up to ~`1e6` or so. You need high
permutations to make claims about stuff rare enough to survive
p-value adjustment as significant, so test on low significance and
then let it run in looking for real significance.

Since we did this once, we can do this multiple times with a loop.
Note this includes `system.time`ing.

```{r,multiple,dependson="sortedGeneList",cache=T,eval=T}
nPermutations <- 1e4

# A list for results holding
GSEAResults <- list()

for (varz in c("PostShift_Mean","PreShift_Mean","Change_Mean")) {
  message(varz)
  message(" took ... ")
  print(system.time(
    {
    this_gene_list <- datar[order(datar[[varz]],decreasing=T),] %>%
      left_join(SystematicToSGDIDmapper,by="Systematic")%>%
      {setNames(.[[varz]],nm=.[["SGDID"]])}%>%na.omit()%>%c()
    GSEAResults[[varz]] <- clusterProfiler::GSEA(
      geneList=this_gene_list,
      TERM2GENE=SGDGOSlim[,c("X6","X3")],
      TERM2NAME=SGDGOTerms[,c("GOID","X2")],
      minGSSize=3,maxGSSize=500,
      pAdjustMethod="fdr",
      pvalueCutoff=0.3,seed="seedy",
      nPerm=nPermutations
      )@result %>% as_tibble() %>% mutate(Variable=varz)
    }
  ))
}
```

Once done, we can print this sucker out as a CSV.

```{r,tableOut,cache=T}
gseGOtable <- bind_rows(GSEAResults) %>% as_tibble %>% 
    # First we bind rows and make it a tibble
  select(Variable,ID,Description,enrichmentScore,p.adjust
    ,setSize,core_enrichment) %>%
    # Then we pick the variables we want
  arrange(Variable,-enrichmentScore) %>%
  mutate(sgdIDz=map(core_enrichment,str_split,"/",simplify=F)) %>%
    # This is weird. It's taking the core_enrichment thing, 
    # splitting it into a vector
  mutate(commonz=map(sgdIDz,function(x){
    # Then it maps each vector into this anonymous function
    tibble(SGD=unlist(x))%>%
      # unlists it out of a list of a vector into a vector
      left_join(SystematicToSGDIDmapper%>%rename(SGD=SGDID),by="SGD")%>%
        # then joins it against the mapper tibble from before
      pull(Common)
        # because then we can pull out a vector of common names
    })) %>%
  select(-sgdIDz,-core_enrichment)%>%
  mutate(commonz=map(commonz,str_replace_na))%>%
    # we have to remove the NA values, because mysteriously this thing
    # has some of those
  mutate(commonz=unlist(map(commonz,str_c,collapse=",")))
    # then collapse with a comma for reading
```

```{r,printOut,cache=T}
write_lines("#
# Supplementary Table
#
# Results of that gene set enrichment analysis. Put a description
# here please.
#
# Key:
# Variable sorted by , GO term ID , GO term , enrichment score , adjusted p-value , size of GO term set , common names"
  ,path="output_table_GSEA.csv")

gseGOtable %>% 
  write_csv(path="output_table_GSEA.csv",append=T)
```

You can also map this to KEGG terms as well. Use 
`clusterProfiler::enrichKEGG` similarly, but be mindful of the 
`keyType` and translation of different identifiers.


```{r}
sessionInfo()
```




